---
title : "Bjørnar's assignment in Gretl"
#format: html
format: pdf
#format: docx
author:
  - name: Henrik Georg Jacobsen
    affiliations:
      - Høgskolen i Østfold
      
toc: true

---

[Bjørnar's assignment](https://hiof.instructure.com/courses/9596/files/folder/Assignment?preview=1695176) 

Inline equations $\theta=2+2$


## Exercise 1 

```
      obs           price          WinterRain         temp        HarvestRain         Age       
 Min.   :1952   Min.   : 495.2   Min.   :376.0   Min.   :14.98   Min.   : 38.0   Min.   : 3.00  
 1st Qu.:1959   1st Qu.: 670.8   1st Qu.:543.5   1st Qu.:16.15   1st Qu.: 88.0   1st Qu.: 9.50  
 Median :1966   Median :1079.8   Median :600.0   Median :16.42   Median :123.0   Median :16.00  
 Mean   :1966   Mean   :1405.8   Mean   :608.4   Mean   :16.48   Mean   :144.8   Mean   :16.19  
 3rd Qu.:1973   3rd Qu.:1707.7   3rd Qu.:705.5   3rd Qu.:17.01   3rd Qu.:185.5   3rd Qu.:22.50  
 Max.   :1980   Max.   :4883.9   Max.   :830.0   Max.   :17.65   Max.   :292.0   Max.   :31.00  
                NA's   :2        NA's   :2       NA's   :2       NA's   :2       NA's   :2      

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

wine <- read.csv("wine.csv")


wine_clean <- na.omit(wine)

ggplot(wine_clean, aes(x = price)) +
  geom_histogram(binwidth = 200, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Wine Prices", x = "Price", y = "Count")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
wine <- read.csv("wine.csv")

ggplot(wine, aes(x = temp, y = price)) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Scatterplot: Wine Price vs Temperature",
    x = "Temperature (°C)",
    y = "Price"
  )
```







## Exercise 2 
*Model 1*
$$
price_i = \beta_0 + \beta_1Age_i + u_i 
$$

a.
```
Model 3: OLS, using observations 1–29
Dependent variable: price

             Coefficient  Std. Error  t-ratio  p-value
const         494.278      402.112     1.229   0.2304
Age           56.318       22.222      2.534   0.0179 **

R-squared: 0.2043
Adjusted R-squared: 0.1728
F-statistic: 6.420
P-value (F): 0.0179
``` 

b.

For each year the wine gets older, its price is expected to increase by 56.318 units.

c.

P-value (0.0179)>α(0.01)

Significance level (α) = 0.01 (1%)

$H_0: \beta = 0$

$H_1: \beta \ne 0$

P-value is higher than 0.01, that means age is not statistically significant.

d.

```{r, echo = FALSE}
wine <- read.csv("wine.csv")

age_min <- min(wine$Age, na.rm = TRUE)
age_mean <- mean(wine$Age, na.rm = TRUE)
age_max <- max(wine$Age, na.rm =TRUE)

model <- lm(price ~ Age, data = wine)

age_values <- data.frame(Age = c(age_min, age_mean, age_max))

predicted <- predict(model, newdata = data.frame(Age = age_values))

data.frame(Age = age_values, Predicted_Price = round(predicted, 2))



```


## Exercise 3 
*Model 2*
$$
price_i = \beta_0 + \beta_1Age_i + \beta_2WinterRain_i  +\beta_3temp_i + \beta_3HarvestRain_i + u_i \\
$$

a.

``` 
Model 4: OLS, using observations 1952-1980 (T = 27)
Missing or incomplete observations dropped: 2
Dependent variable: price

                coefficient    std. error    t-ratio   p-value 
  -------------------------------------------------------------
  const         −15509,0       3379,87       −4,589    0,0001   ***
  Age               39,2126      14,3490      2,733    0,0121   **
  WinterRain         2,75098      0,965119    2,850    0,0093   ***
  temp             930,787      190,557       4,885    6,97e-05 ***
  HarvestRain       −5,04694      1,61682    −3,122    0,0050   ***

Mean dependent var   1405,800   S.D. dependent var   1027,226
Sum squared resid     7238994   S.E. of regression   573,6246
R-squared            0,736141   Adjusted R-squared   0,688166
F(4, 22)             15,34443   P-value(F)           3,93e-06
Log-likelihood      −207,0499   Akaike criterion     424,0999
Schwarz criterion    430,5791   Hannan-Quinn         426,0265
```

b.

### Age:

Coefficient = 39.21

P-value = 0.0121

Wine price increases by 39.21 each year.

### WinterRain:

Coefficient = 2.75

P-value = 0.0093

Wine price increases by 2.75 each mm winter rainfall.

### Temperature:

Coefficient = 930.79

P-value = 0.00007

Wine price increases by 930.79 every 1 degree.

### HarvestRain

Coefficient = -5.05

P-value = 0.0050

Wine price decreases by 5.05 each additional 1 mm HarvestRain

c.

Adding more variables improves the model to explain the wine prices.

## Exercise 4 
*Model 3*
$$
price_i = \beta_0 + \beta_1Dheavyraint_i + \beta_2tempt_i + \beta_3temp_i · Dheavyrain_i + u_i
$$

a.

```
Model 2: OLS, using observations 1952-1980 (T = 27)
Missing or incomplete observations dropped: 2
Dependent variable: price

                    coefficient   std. error   t-ratio   p-value
  --------------------------------------------------------------
  const             −16289,5       4395,18     −3,706    0,0012  ***
  Dheavyrain         11634,1       8990,38      1,294    0,2085 
  temp                1082,95       266,391     4,065    0,0005  ***
  temp_Dheavyrain     −756,902      546,064    −1,386    0,1790 

Mean dependent var   1405,800   S.D. dependent var   1027,226
Sum squared resid    14038659   S.E. of regression   781,2660
R-squared            0,488295   Adjusted R-squared   0,421550
F(3, 23)             7,315911   P-value(F)           0,001294
Log-likelihood      −215,9914   Akaike criterion     439,9829
Schwarz criterion    445,1662   Hannan-Quinn         441,5242

Excluding the constant, p-value was highest for variable 6 (Dheavyrain)
```

b.

Temp * Dheavyrain captures whether the impact of temperature on price depends on harvest rainfall. 
Dheavyrain captures the shift in wine price when harvest rainfall is very high, regardless of temperature.




## Exercise 5 
a.

```
| Model     | Actual Price | Predicted Price  | Residual |
|-----------|--------------|------------------|----------|
| Model 1   | 4883.90      | 1733.28          | 3150.62  |
| Model 2   | 4883.90      | 3578.80          | 1305.10  |
| Model 3   | 4883.90      | 2481.70          | 2402.20  |
```


b.

Model 2 is the best model for predicting Bordeaux wine price. 

It hast multiple significant variables and is a better fit. 










